---
title: Visualise and investigate protein phosphorylation in non-smoking females by
  tumor stage
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

## 1. Introduction
According to the paper, differentially regulated proteins are strongly separated among three proteomic subtypes (Subtype 1, Subtype 2, Subtype 3). Subtype 1 is mostly comprised of late stage tumors (≥II), subtype 2 mostly of early stage patients(IA and IB) and subtype 3 represents over-representation of early stage patients(IA). The overall results showed that subtype 3 had the highest phosphorylation of proteins. However, as reported by the proteome-based classification, subtype 1 can be further divided into stage IA late-like(IIA) and stage IB late-like(IIB). Therefore, this paper revealed 5 distinct stage tumors which are: ≥II, IA, IB, IIA and IIB. Instead of comparing protein phosphorylation amongst different subtypes, whether the correlation of protein phosphorylation can be visualised by tumor stage was considered. Due to the vast number of proteins, only proteins with 0 ANOVA q value was utilised. ANOVA q value, also known as FDR is defined as a p-value that has been adjusted for the proportion of false positives from the test. In other words, if the q value is 0, 0% of the compounds will result in false positives.

## 2. Setting the data
Download the necessary packages
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(ggside)
library(forcats)
library(ggridges)
library(cowplot)
library(ggpubr)
```

Import the excel file and specify the sheet that is needed.
```{r}
library(readxl)
readxl::excel_sheets('1-s2.0-S0092867420307431-mmc6.xlsx')
a = read_excel('1-s2.0-S0092867420307431-mmc6.xlsx', sheet=2, col_names=TRUE)
```

Change the header of the column names so that there is no space and break line.
```{r}
colnames(a) <- gsub(" ","_",colnames(a))
colnames(a) <- gsub("\r?\n|\r","_",colnames(a))
```

## 3. Manipulate the data frame and plot the first visualisation
In order to select a specific range in the ```ANOVA__q-value', it should first be converted to numeric values. After applying the range, change the class back into a character for rearrangment of the dataframe.
```{r}
a$`ANOVA__q-value` <- as.numeric(a$`ANOVA__q-value`) 
a <- subset(a, `ANOVA__q-value` = 0 | is.na(`ANOVA__q-value`))
a$`ANOVA__q-value` <- as.character(a$`ANOVA__q-value`) 
```

Switch the column and row headers to filter out and select specific values. Then, readjust the new column headers so that it does not have any spaces. 
```{r}
b <- a %>%
   tibble::rownames_to_column() %>%  
   pivot_longer(-rowname) %>% 
   pivot_wider(names_from=rowname, values_from=value)
colnames(b) <-b[1,]
b <- b[-1,]
colnames(b)<-gsub(" ","_",colnames(b))
as.data.frame(b)
```
Filter female non-smokers, and select only the columns that are used in the ggplot. Switch the column and row back again into a longer format so that is possible to plot a heatmap.
```{r}
c <- b %>% filter(SmokingStatus=="no", Gender=="female") %>% 
  select(OverallStaging, id, 81:270) %>% 
  slice(5:n()) %>% 
  pivot_longer(cols= !c(OverallStaging, id),
               names_to= "Protein", 
               values_to="Log2T/N")
c$`Log2T/N`<- as.numeric(c$`Log2T/N`)
```

Plot the heatmap.
```{r, fig.height=25, fig.width=25}
c %>% 
  group_by(OverallStaging) %>% 
  arrange(OverallStaging) %>%
  ggplot(aes(x=fct_inorder(id), y=Protein, fill=`Log2T/N`)) +
  theme(text=element_text(size=30), 
        axis.text.x = element_text(angle =90, size=15, vjust = 0.4), 
        axis.text.y = element_text(size=6)) +
  scale_fill_gradient2(midpoint=0,
                       limit=c(-4.4,7.3),
                       name='Log2T/N',
                       low="blue",
                       mid="white",
                       high="red",
                       na.value = "black") +
  geom_tile() + 
  guides(fill=guide_colourbar(barwidth=2.5, barheight = 30)) +
    geom_xsidetile(aes(y=OverallStaging, xfill=OverallStaging)) +
    scale_xfill_manual(values=c("red","orange","yellow","green","blue")) +
  labs(title="Differentially regulated proteins by tumor stage in non-smoking females",
       subtitle = "Proteins with ANOVA qvalue of 0",
  x="Patient ID", 
  y="Protein")
```

From the figure above, it is evident that bumber of non-smoking female patients are highest in stage IA, whereas there is only one patient in tumor stage IIB. As seen on the heatmap, both highest and lowest phosphorylation of proteins exist in tumor stage IA. In order to take a closer look and visualise the difference of protein phosphorylation among tumor stage, 5 proteins with highest phosphorylation (P15088, P23946, P55083, Q9BU40, P08311)and lowest phosphorylation(O75940, O95171, O15541, Q9UKY7, Q16625) were chosen resepectively. As expected, 10 of these proteins all existed in tumor stage IA. 

## 4. Plot the second visualisation
Select 10 specific proteins with highest(P15088, P23946, P55083, Q9BU40, P08311) and lowest(O75940, O95171, O15541, Q9UKY7, Q16625) phosphorylation and turn the data frame into a longer format for it to be possible to plot ridgelines.
```{r}
d <- b %>% filter(SmokingStatus=="no", Gender=="female") %>% select(P15088, P23946, P55083, Q9BU40, P08311, OverallStaging) %>% group_by(OverallStaging) %>% pivot_longer(col=P15088:P08311, names_to="H_Protein",values_to="Log2T/N")
d$`Log2T/N` <- as.numeric(d$`Log2T/N`)
e <- b %>% filter(SmokingStatus=="no", Gender=="female") %>% select(O75940, O95171, O15541, Q9UKY7, Q16625, OverallStaging) %>% group_by(OverallStaging) %>% pivot_longer(col=O75940:Q16625, names_to="L_Protein",values_to="Log2T/N")
e$`Log2T/N` <- as.numeric(e$`Log2T/N`)
```


Plot ridgelines
```{r}
HP_stage <- d %>%
  ggplot(aes(x=`Log2T/N`, y= fct_reorder(H_Protein,`Log2T/N`), fill=stat(x))) +
    geom_density_ridges_gradient() +
    geom_point(aes(colour=`Log2T/N`)) +
    scale_color_gradient2(high = "blue", mid = "white", low = "red") +
  guides(fill = "none") +
  scale_fill_gradient2(high = "blue", mid = "white", low = "red") +
  scale_x_continuous(limits = c(-5, 7)) +
   coord_cartesian(clip = "off") +
  labs(x = 'Log2T/N',
       y = 'Protein',
       fill = 'Log2T/N',
       title = "Highest Expressed Protein in each tumor stage") +
      facet_wrap(.~OverallStaging) 

LP_stage<- e %>%
  ggplot(aes(x=`Log2T/N`, y= fct_reorder(L_Protein,`Log2T/N`), fill=stat(x))) +
    geom_density_ridges_gradient() +
    geom_point(aes(colour=`Log2T/N`)) +
    scale_color_gradient2(high = "blue", mid = "white", low = "red") +
  guides(fill = "none") +
  scale_fill_gradient2(high = "blue", mid = "white", low = "red") +
  scale_x_continuous(limits = c(-4, 4)) +
   coord_cartesian(clip = "off") +
  labs(x = 'Log2T/N',
       y = 'Protein',
       fill = 'Log2T/N',
       title = "Lowest Expressed Protein in each tumor stage") +
      facet_wrap(.~OverallStaging)
```

```{r,fig.height=4, fig.width=12}
ggarrange(HP_stage,LP_stage, labels = c("A","B"))
```
Plotting ridgelines with 5 proteins from each high and low phosphorylation clearly showed the difference of phosphorylation between tumor stages. Both highest and lowest phosphorylated proteins were in tumor stage IA, which can be seen by the density of the colour compared with other tumor stages. Then, tumor stage IB always came next for high and low phosphorylation of proteins. Interestingly, when proteins were highly or lowly phosphorylated in tumor stage IA and IB, they were in contrast lowly and highly phosphorylated repectively in tumor stages IIA and ≥II. However, there is only one patient in tumor stage IIB, which is why it is impossible to draw a ridgeline, thus phosphorylation values of each proteins are shown as a dot. In stage IIB, selected proteins for each highest and lowest phosphorylation strongly shows that all these proteins are in correlation as none of the values were oppositely expressed.

## 5. Discussion
After drawing the heatmap and plotting ridgelines with selected highly and lowly phosphorylated proteins, the results showed a clear difference between tumor stage IA, IB and IIA and ≥II. Research showed that 3 out of 5 highest phosphorylated proteins (P15088, P23946, P08311) in tumor stage IA had a function of angiotensin maturation. With careful consideration, as tumor stage IA and IB is early stage, protein with angiotensin maturation would be highly phosphorylated in order for cancer cells to develop. This is only based on 3 proteins that are selected within small range, therefore it may not always be the case which means that this can be a further investigation later on. 
However, there is also a limitation to these plots. In order to minimise the number of proteins, values with 0 ANOVA qvalues were chosen. This ANOVA qvalue is calculated within the different subtypes, which means that it might not be as accurate when used to compare between different tumor stages. Even though, this is the case, due to the fact that the subtypes were mostly divided by stages, it was thought to be possible to see a correlation. 